\begin{abstract}
\label{abstract}
    旅行商问题（TSP）是最经典的组合优化问题之一，其已广泛存在于大部分行业，包括但不限于交通运输、网络通信和钻孔电路板等领域。同时，多目标旅行商问题（MOTSP）具是有多个冲突的目标的多目标组合优化问题，设计相应的多目标组合优化算法来解决这类问题具有重要的实际意义。本文围绕MOTSP和基于分解的多目标组合优化算法进行研究，主要研究内容包括以下两个部分：
    \par
    第一，TSP拥有巨大的搜索空间会导致局部搜索算法的搜索时间过长、易陷入局部最优。针对这一问题，可以通过对具有显性结构化搜索空间进行剪枝预处理，从而达到提高算法的运行效率的目的。本文提出了基于最小生成树和欧拉回路的邻域结构生成算法，并通过局部搜索技术对这些邻域结构进行搜索，以此提升问题的优化效率。通过实验对比，可以证实基于最小生成树和欧拉回路的邻域结构生成算法生成邻域结构的高效率及其邻域结构的高质量。
    \par
    第二，基于上述TSP的邻域结构，结合进化迁移中不同领域信息迁移这一特征，提出了基于邻域结构迁移的多目标旅行商问题求解算法。并在邻域结构迁移中构建了两目标相似度模型来引导算法进行邻域结构正迁移，同时构造了一个多臂老虎机模型来学习选择邻域结构的知识，以便进行更有效的局部搜索。最后，通过实验能够表明算法在各种测试用例上获得的非支配解集不仅分布性好，而且收敛性也优异，并且验证了邻域结构迁移中的两目标相似度模型和多臂老虎机模型的有效性。

    \end{abstract}
    \keywords{多目标优化, 旅行商问题, 局部搜索, 进化迁移优化, 邻域结构, 邻域结构迁移}

\begin{abstractEn}
\label{abstractEn} 
    The traveling salesman problem (TSP) is one of the most classic combinatorial optimization problems. It has been widely used in most industries, including but not limited to transportation, network communications, and drilling circuit boards. At the same time, the multi-objective traveling salesman problem (MOTSP) is a multi-objective combinatorial optimization problem with multiple conflicting objectives. It is of great practical significance to design a corresponding multi-objective combinatorial optimization algorithm to solve this type of problem. This thesis focuses on MOTSP ​​and decomposition-based multi-objective combinatorial optimization algorithms.
    The main research content of this thesis includes the following two parts:
    \par
    First, TSP has a huge search space, which will cause the search time of the local search algorithm to be too long, and it is easy to fall into the local optimum. To solve this problem, the pruning preprocessing can be performed on the explicit structured search space, so as to achieve the goal of improving the efficiency of the algorithm. This thesis proposes a neighbor structure generation algorithm based on minimum spanning tree and Euler circuit, and searches these neighborstructures through local search technology to improve the optimization efficiency of the problem. Through experimental comparison, it can be confirmed that the neighbor structure generation algorithm based on minimum spanning tree and Euler circuit has the high efficiency of generating neighbor structure and the high quality of neighbor structure.
    \par
    Second, based on the above-mentioned neighbor structure of TSP, a multi-objective TSP evolutionary algorithm based on neighbor structure transfer is proposed. To guide positive knowledge transfer, a bi-objective similarity model is proposed to generate a neighbor structure candidate set for each subproblem. At the same time, a multi-armed bandit model is constructed to learn the knowledge of which neighbor structure to use for more efficient local search on the run. Finally, through experiments, it can be shown that the non-dominated solution set obtained by the algorithm on various test cases is not only well distributed, but also excellent in convergence. It also verifies the bi-objective similarity model and the multi-armed bandit model in the neighborhood structure migration. Effectiveness.
\end{abstractEn}
\keywordsEn{Multi-objective Optimization, Traveling Salesman Problem, Local Search, Evolutionary Transfer Optimization, Neighbor Structure, Neighborhood Structure Transfer}
